A biometric analysis of infant mortality and temperature, northern Sweden 1895--1950

### 2022-02-24 paper.Rmd
final commit

---
title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 4
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
- \usepackage[tablesonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895-1950 is studied in a biometric analysis setting. More precisely, the effect of climate and weather, measured by temperature, average and extremes, on infant mortality is investigated. It turns out that climate (average) is more important than weather (extremes), low average temperatures are more important than temporary dips in temperature, but effects are different in neonatal and postneonatal settings."
---

@@

```{r readdata}
births <- readRDS("data/births4.rds")
infdat  <- readRDS("data/infdat4.rds")
```
# Introduction

The impact of ambient temperature variations on infant mortality is studied for 
two northern Sweden areas, the Umeå coastal region and Skellefteå coastal and inland regions, during the 
first half of the twentieth century. Two recent papers [@junkka; @karlsson] studied 
neonatal mortality and temperature variations in this geographical area during the 
years 1880--1950. Climate and mortality in general is a research area that has 
generated great interest over the last years, see @tbgb10 for an example of our own efforts.

The effect of seasonal variation and the occurrence of extreme monthly temperatures
is studied and interacted with social class and time period. Studies are performed 
separately for neonatal  and postneonatal mortality, and for winter and summer seasons.

@@

```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
pneo <- age.window(infdat, c(28 / 365, 1))
tpneo <- toTpch(Surv(enter, exit, event) ~ week, data = pneo, cuts = c(28 / 365, 1))
y <- tpneo$event / tpneo$exposure
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "O/E rate", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, at = c(0, 0.02, 0.04, 0.06), las = 1)
abline(h = c(0.03, 0.04, 0.05, 0.06, 0.07), lty = 3)
abline(h = 0)
box()
```

@@

```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4, cache = TRUE}
ndat <- age.window(infdat, c(0, 28 / 365))
tnda <- toTpch(Surv(enter, exit, event) ~ year, data = ndat, cuts = c(0, 28 / 365))
y <- tnda$event / tnda$exposure
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "O/E rate", lwd = 1.5, cex = 0.7)
abline(h = 0)
abline(h = c(0.2, 0.4, 0.6), lty = 3)
```

```{r pntrend, fig.cap = "Crude postneonatal mortality by year, Umeå-Skellefteå 1895--1950.", fig.height = 4, cache = TRUE}
pndat <- age.window(infdat, c(28 / 365, 1))
tpnda <- toTpch(Surv(enter, exit, event) ~ year, data = pndat, 
                cuts = c(28 / 365, 1))
y <- tpnda$event / tpnda$exposure
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "O/E rate", lwd = 1.5, cex = 0.7)
abline(h = 0)
abline(h = c(0.2, 0.4, 0.6), lty = 3)
```

@@

The two data sets, mortality and weather, are combined into one by treating temperature data 
as a communal covariate and incorporate it as such in the mortality data set. The 
function *make.communal* in the **R** [@cran] package *eha* [@eha; @ehar2] is used 
for that purpose.  Resulting data drame is partly shown in Table \@ref(tab:readdata3).

```{r readdata3}
levels(infdat$subreg) <- c("ume", "ske", "inland")
##infdat$excessTemp <- cut(infdat$extemp, c(-16, -2, 2, 9), include.lowest = TRUE) # NOTE!!!
##infdat$excessTemp <- relevel(infdat$excessTemp, ref = "(-2,2]")
##infdat$lowTemp <- (infdat$extemp + infdat$extemp.1) < -10  # CHECK THIS!!
##infdat$highTemp <- infdat$extemp + infdat$extemp.1 > 10
##infdat$lowTemp <- with(infdat, (extemp < -3) & (extemp.1 < -3))
infdat$lowTemp <- (with(infdat, cold + cold.1 < -10))
infdat$highTemp <- with(infdat, (extemp > 3) & (extemp.1 > 3))
tbl(head(infdat[, c("enter", "exit", "event", "extemp", "extemp.1", "emeantemp", "week", "year")]),
    caption = "Data with communal covariates.", fs = 9)
```

@@

It turns out that extremely low temperature (`lowTemp`) is bad during all seasons except summer,
and extremely high temperature (`highTemp`) is bad during summer, but good otherwise.
So we group season into two categories, *summer* and *not summer*.
In each case separate analyses for neonatal and postneonatal mortality are performed.

@@

```{r prepare}
source("R/ltx2.R")
infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
##infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
##levels(infdat$subreg) <- c("ume", "ske")
##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
##infdat$season <- as.factor(infdat$season)
## New:
infdat$summer <- infdat$week %in% 14:39
infdat$emeantemp <- round(infdat$emeantemp)
```

@@

The results regarding neonatal mortality is much in accordance with the results
found by @junkka. However, they used temperature in a "hockey-stick" regression with a breakpoint at
14.5 degrees Celsius and a negative slope (decreasing risk) to the left and a positive slope
(increasing risk) to the right. Instead, we are using the average weekly temperature
for the 52 weeks of a year, for each week averaging over all the years in the study, as
our "reference points" ("climate"), adding deviances up and down ("weather") as
"short-term temperature stress". This is similar to the way prices and mortality were related
in for instance @genus11, that is, a time series split into long time trend and short
term variation.

Comments on other candidate covariates:

*Birth month* is left out in the analyses despite that fact that
it is an important factor in neonatal mortality. However, we include *time of year* in terms of 
*winter* and *summer* as a time-varying covariate, and in the neonatal case it will coincide to a
great extent with birth period. For the postneonal case the situation is different, but it turns out that
for those infants who have survived the first month of life, birth month does not matter much.
We separate the investigation into two parts, *neonatal* and *postneonatal* mortality. But first,
a joint analysis.

*Socioeconomic status* is divided into two factor covariates: *socBranch* and *socStatus*. The latter should be 
seen relative to actual *socBranch*. We have information on whether the infant was *illegitimate* (mother unmarried),
but we incorporate those cases in the category *none* of *socBranch* and *unknown* as category of *socStatus*.

```{r managedata}
if (!(file.exists("data/inftab.rds") && file.exists("data/tabneo.rds") && file.exists("data/tabpost.rds"))){
    neo <- age.window(infdat, c(0, 28 / 365))
    neo$ageIvl <- "neo"
    pneo <- age.window(infdat, c(28/365, 1))
    pneo$ageIvl <- "postneo"
    all <- rbind(neo, pneo)
    all$ageIvl <- factor(all$ageIvl)
    ##all$enter <- round(all$enter * 365)
    ##all$exit <- round(all$exit * 365)
    ##fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (extemp + emeantemp + lowTemp + highTemp + socBranch + ##socStatus + sex + period),
    ##               data = all)
    ##dr.all <- drop1(fit.all, test = "Chisq")

    inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socst + urban + parity + 
                         extemp + extemp.1 + emeantemp + summer + ageIvl, data = all,
                     cuts = c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365) / 365)
    tabneo <- inftab[inftab$ageIvl == "neo", ]
    tabpost <- inftab[inftab$ageIvl == "postneo", ]
    tabpost$age <- factor(tabpost$age)
    tabneo$age <- factor(tabneo$age)
    saveRDS(tabneo, file = "data/tabneo.rds")
    saveRDS(tabpost, file = "data/tabpost.rds")
    saveRDS(inftab, file = "data/inftab.rds")
}else{
    inftab <- readRDS("data/inftab.rds") #The rest are read later...
}
```

```{r joint, cache = TRUE}
fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * 
                       (extemp + extemp.1 + emeantemp + socst + period + urban), 
                   data = inftab, time = age)
dr.all <- drop1(fit.all, test = "Chisq")
##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socst, 
  ##               data = tabneo, time = age))
##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socst, 
  ##               data = tabpost, time = age))
```

Covariates may affect neonatal and postneonatal mortality differently, and one way to 
investigate that is to stratify the infant mortality data into two age intervals, one from
birth to one month of age and the other from one month to one year of age. Then the
interaction between the stratum variable and other variables of interest is investigated, 
with this result:

```{r shownonprop, cache = FALSE}
dr.all
##tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
```

We can see that the variables *emeantemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case. In terms of 
*infant mortality*, we can say that these caovariates have non-proportional effects.

@@

The analyses are split into two parts by season, *winter* is one, and *summer* the other.
Summer covers the months April to September, and winter the rest of the months.

```{r formneo}
formneo <- as.formula("oe(event, exposure) ~  extemp + extemp.1 + emeantemp + socst + period + urban +
                     subreg + sex + parity")
tabneo <- readRDS("data/tabneo.rds")
```

@@

This period refers to the months *October to March*. A Cox regression involves as
interesting variables *highTemp*, an indicator of temperature at least four degrees
above the expected for at least two weeks in a row, *emeantemp* the expected temperature
the actual weeek, and *extemp* the *excess temperature* the actual week.

```{r neoanalwp, results = 'asis', cache = TRUE}

##library(survival)
winter <- tabneo[!tabneo$summer, ]
##winter$season <- factor(winter$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.neow <- tpchreg(formneo, data = winter, time = age)
dr <- drop1(fit.neow, test = "Chisq")
```

```{r neoanalw, results = "asis", cache = FALSE}
ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March.", 
     label = "tab:neoanalw", keep = NULL)
```

#### Temperature and sex

We check the interaction between temperature and sex and period.

```{r neosexper}
fit <- tpchreg(oe(event, exposure) ~ sex * (extemp + extemp.1 + emeantemp), 
               time = age, data = winter)
dr <- drop1(fit, test = "Chisq")
dr
```

Obviously no temperature and sex interactions

@@

This period refers to the months *April to September*. A Cox regression involves as
interesting variables *highTemp*, an indicator of temperature at least four degrees
above the expected for at least two weeks in a row, *emeantemp* the expected temperature
the actual weeek, and *extemp* the *excess temperature* the actual week.

>>>>

---
title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895--1950"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 4
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
- \usepackage[tablesonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895--1950 is studied in a biometric analysis setting. More precisely, the effect of climate and weather, measured by temperature, average and extremes, on infant mortality is investigated. The main question is to what extent temperature, directly or indirectly via humidity, affect the survival of infants and to what extent it depends on the resources of the family. It turns out that the normal seasonality (the average temperature) is generally more important than the extremes (the deviations), low average temperatures are more important than temporary dips in temperature, but effects are different in neonatal and postneonatal settings. It is at normal temperatures that we expect the resources of the family to be of most importance, and this is also what we find. In extremely cold winters, also the children of the higher classes suffered from increased postneonatal mortality."
---

@@

```{r readdata}
births <- readRDS("data/births4.rds")
infdat <- readRDS("data/infdat4.rds")
infdat$marginTemp <- round(infdat$extemp.1)
infdat$expTemp <- round(infdat$emeantemp.1)
```
# Introduction

The impact of ambient temperature variations on infant mortality is studied for 
two northern Sweden areas, the Umeå coastal region and Skellefteå coastal and inland regions, during the 
first half of the twentieth century. Two recent papers [@junkka; @karlsson] studied 
neonatal mortality and temperature variations in this geographical area during the 
years 1880--1950.
@rock2014 studies associations between temperature and age-stratified mortality,
Skellefteå 1749--1859. Effects of temperature on mortality in general is a research area that has generated great interest over the last years, partly due to new weather conditions related to climate change.

In our previous study, we analysed the role of climate for mortality in different ages in southern Sweden 1766--1860 combining time-series data on temperature and longitudinal individual level data
[@tbgb10]. We found that low winter temperatures led to an increase in mortality among adults but not children in ages 1-15 years [@tbgb10]. To understand how hard winters and warm summers affected different parts of the population, we used methods similar to the ones we have developed for our analyses of the impact of food prices. This means that the focus was on annual variations on temperature and mortality. The main focus was on demographic vulnerability to changes in external factors, such as food process and temperature, as a measurement of living standards. Well-off groups could afford buying food and firewood even of prices went up, while those living close to their margins suffered to such an extent that family members might die [@tb04]. Still, child mortality was fairly equal between social classes until the latter part of the nineteenth century to widen in the beginning of the twentieth century, for both post-neonatal infant and child mortality [@dk21]. For neonatal mortality, no systematic class differences existed in any period. While infant was equal between social classes up until the end of the nineteenth century, it differed between urban and rural areas. In fact, in 1850, infant mortality in urban areas was 50 percent higher than in rural areas and it was not until the 1920s that mortality rates converged [@scb69].

Periods with low temperature might not only cause problems of keeping body heat, indoors and outdoors, but also lead to an increase in the spread of virus. The reason is that low temperature is associated with low humidity. A study of England and Wales in 1921--27, for example, shows the close connections between seasonality in temperature, humidity, and diagnosed cases of smallpox, which was a mild disease by then [@rl28]. This means that you are more likely to be infected by viral diseases, such as measles, whooping-cough, and influenza, in low temperatures. Since the outcome of such diseases are not dependent on the nutritional status of the host [@rr83], we anticipate that they will not have a social gradient. Other diseases that are nutritionally dependent, such as pneumonia, TB, and gastro-intestinal diseases, might, however, have such a gradient. It means that low winter temperatures might give rise to a social gradient in mortality, but it may as well be socially neutral.

In this study, we differentiate between neonatal and postneonatal mortality in an effort to understand the role of resources to protect infants rom both normal and extreme ambient temperatures. We analyse the effects of seasonal variation and the occurrence of extreme weekly temperatures and its interaction with social class and time period. Analyses are performed separately for neonatal and post neonatal mortality, and for winter and summer seasons. The reason is that mortality in the first four weeks of life is associate with different factors than mortality later in infancy. For this reason, Bourgeois-Pichat differentiate between endogenous and exogenous factors, arguing that mortality shortly after birth is due to the maternal health, fetal factors, and the delivery itself, whereas it afterwards is due to exposure to external factors following the birth-weight development [@bp51a;@bp51b]. The main question is to what extent temperature, directly or indirectly via humidity, affect the survival of infants and to what extent it depends on the resources of the family.

@@

```{r pneodata, fig.cap="Crude postneonatal mortality by week of year, cache=FALSE, Umeå/Skellefteå 1895--1950.", fig.height=3}
pneo <- age.window(infdat, c(28 / 365, 1))
tpneo <- toTpch(Surv(enter, exit, event) ~ week, data = pneo, cuts = c(28 / 365, 1))
y <- tpneo$event / tpneo$exposure
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "O/E rate", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, at = c(0, 0.02, 0.04, 0.06), las = 1)
abline(h = c(0.03, 0.04, 0.05, 0.06, 0.07), lty = 3)
abline(h = 0)
box()
```

@@

```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
ndat <- age.window(infdat, c(0, 28 / 365))
tnda <- toTpch(Surv(enter, exit, event) ~ year, data = ndat, cuts = c(0, 28 / 365))
y <- tnda$event / tnda$exposure
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "O/E rate", lwd = 1.5, cex = 0.7)
abline(h = 0)
abline(h = c(0.2, 0.4, 0.6), lty = 3)
```

```{r pntrend, fig.cap = "Crude postneonatal mortality by year, Umeå-Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
pndat <- age.window(infdat, c(28 / 365, 1))
tpnda <- toTpch(Surv(enter, exit, event) ~ year, data = pndat, 
                cuts = c(28 / 365, 1))
y <- tpnda$event / tpnda$exposure
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "O/E rate", lwd = 1.5, cex = 0.7)
abline(h = 0)
abline(h = c(0.2, 0.4, 0.6), lty = 3)
```

@@

The two data sets, mortality and weather, are combined into one by treating temperature data 
as a communal covariate and incorporate it as such in the mortality data set. The 
function *make.communal* in the **R** [@cran] package *eha* [@eha; @ehar2] is used 
for that purpose.  Resulting data frame is partly shown in Table \@ref(tab:readdata3).

```{r readdata3}
levels(infdat$subreg) <- c("ume", "ske", "inland")
##infdat$excessTemp <- cut(infdat$extemp, c(-16, -2, 2, 9), include.lowest = TRUE) # NOTE!!!
##infdat$excessTemp <- relevel(infdat$excessTemp, ref = "(-2,2]")
##infdat$lowTemp <- (infdat$extemp + infdat$extemp.1) < -10  # CHECK THIS!!
##infdat$highTemp <- infdat$extemp + infdat$extemp.1 > 10
##infdat$lowTemp <- with(infdat, (extemp < -3) & (extemp.1 < -3))
infdat$lowTemp <- (with(infdat, cold + cold.1 < -10))
infdat$highTemp <- with(infdat, (extemp > 3) & (extemp.1 > 3))
##infdat$weather <- infdat$extemp.1
##infdat$climate <- round(infdat$emeantemp)
tbl(head(infdat[, c("enter", "exit", "event", "marginTemp", "expTemp", "week", "year")]),
    caption = "Data with communal covariates.", fs = 11)
```

We will use the temperature variables *meantemp* and *emeantemp* under the names *marginTemp* 
and *expTemp*, respectively, lagged one week. That is, we assume that the effect of 
temperatures on mortality show itself the following week.

@@

It turns out that extremely low temperature is bad during all seasons,
and extremely high temperature is bad during summer, but good otherwise.
So we group season into two categories, *summer* and *winter*.
In each case separate analyses for neonatal and postneonatal mortality are performed.

@@

```{r prepare}
source("R/ltx2.R")
infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
##infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
##levels(infdat$subreg) <- c("ume", "ske")
##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
##infdat$season <- as.factor(infdat$season)
## New:
infdat$summer <- infdat$week %in% 14:39
##infdat$eexpTemp <- round(infdat$emeantemp)
```

## Variables

These are the covariates considered in our analyses. Central are the two first, 
already discussed, but of interest is also their effect (on mortality) covariation,
primarily with *social status* (`socst`).

*   **expTemp** (time varying) expected temperature given week number.
*   **marginTemp** (time varying) Observed week temperature minus expTemp.
*   **socst** Social status:
     + *highWhiteC* Upper white collar
     + *farmer*
     + *lowWhiteC* Lower white collar
     + *qualWorker* Qualified worker
     + *worker* Unqualified worker
     + *none* No notation (for instance unmarried mother)
*   **sex** Boy or girl
*   **urban** Small town (Umeå or Skellefteå) or not (logical)
*   **parity** Birth order
*   **year**  (time varying)

Comments on other candidate covariates: 
*Birth month* is left out in the analyses despite that fact that
it is an important factor in neonatal mortality. However, we include *time of year* in terms of 
*winter* and *summer* as a time-varying covariate, and in the neonatal case it will coincide to a
great extent with birth period. For the postneonal case the situation is different, but it turns out that
for those infants who have survived the first month of life, birth month does not matter much.
We separate the investigation into two parts, *neonatal* and *postneonatal* mortality. But first,
a joint analysis.

 We have information on whether the infant was *illegitimate* (mother unmarried);
those cases are incorporated in the category *none* of **socst**.

@@

The results regarding neonatal mortality is much in accordance with the results
found by @junkka. However, they used temperature in a "hockey-stick" regression with a breakpoint at
14.5 degrees Celsius and a negative slope (decreasing risk) to the left and a positive slope
(increasing risk) to the right. Instead, we are using the average weekly temperature
for the 52 weeks of a year, for each week averaging over all the years in the study, as
our "reference points" ("expTemp"), adding deviances up and down ("marginTemp") as
"short-term temperature stress". This is similar to the way prices and mortality were related
in for instance @genus11, that is, a time series split into long time trend and short
term variation.

```{r managedata}
neo <- age.window(infdat, c(0, 28 / 365))
 pneo <- age.window(infdat, c(28/365, 1))
```

```{r nonenow, eval=FALSE}
if (!(file.exists("data/inftab.rds") && file.exists("data/tabneo.rds") && file.exists("data/tabpost.rds"))){

    neo$ageIvl <- "neo"

    pneo$ageIvl <- "postneo"
    all <- rbind(neo, pneo)
    all$ageIvl <- factor(all$ageIvl)
    ##all$enter <- round(all$enter * 365)
    ##all$exit <- round(all$exit * 365)
    ##fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (extemp + emeantemp + lowTemp + highTemp + socBranch + ##socStatus + sex + period),
    ##               data = all)
    ##dr.all <- drop1(fit.all, test = "Chisq")

    inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socst + urban + parity +  marginTemp + expTemp + summer + ageIvl, data = all,
                     cuts = c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365) / 365)
    tabneo <- inftab[inftab$ageIvl == "neo", ]
    tabpost <- inftab[inftab$ageIvl == "postneo", ]
    tabpost$age <- factor(tabpost$age)
    tabneo$age <- factor(tabneo$age)
    saveRDS(tabneo, file = "data/tabneo.rds")
    saveRDS(tabpost, file = "data/tabpost.rds")
    saveRDS(inftab, file = "data/inftab.rds")
}else{
    inftab <- readRDS("data/inftab.rds") #The rest are read later...
}
```

```{r joint, cache=FALSE}
if (!file.exists("Rdata/dr.all.rds")){
    fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * 
                           (extemp.1 + emeantemp.1 + socst + period + urban), 
                       data = inftab, time = age)
    dr.all <- drop1(fit.all, test = "Chisq")
    saveRDS(dr.all, file = "Rdata/dr.all.rds")
}
##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socst, 
  ##               data = tabneo, time = age))
##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socst, 
  ##               data = tabpost, time = age))
```

Covariates may affect neonatal and postneonatal mortality differently, and one way to 
investigate that is to stratify the infant mortality data into two age intervals, one from
birth to one month of age and the other from one month to one year of age. Then the
interaction between the stratum variable and other variables of interest is investigated, 
with the result shown in Table \@ref(tab:shownonprop).

```{r shownonprop, cache = FALSE, results='asis'}
dr.all <- readRDS("Rdata/dr.all.rds")
rownames(dr.all)[2] <- "strata(ageIvl):marginTemp"
rownames(dr.all)[3] <- "strata(ageIvl):expTemp"

tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
```

We can see that the variables *expTemp*, *socst* and *period* all have different effects 
on mortality in the neonatal case compared to the postneonatal case. In terms of 
*infant mortality*, we can say that these covariates have *strong non-proportional* effects.
The response to *marginTemp* variations are probably also different in the neonatal and 
postneonatal periods ($p$-value = 0.09).

Figure \@ref(fig:ph) shows the crude infant mortality, calculated over the complete sample.

```{r ph, fig.cap = "Crude infant mortality."}
##daycuts <- c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365)
##daycuts <- c(0, 3, 7, 28, 56, 140, 183, 250, 365)
b1 <- births
b1$enter <- b1$enter / 365
b1$exit <- b1$exit / 365
daycuts <- c(seq(0, 28, by = 2), seq(49, 245, by = 28), 285, 324, 365)
bb <- toTpch(Surv(enter, exit, event) ~ 1, data = b1, cuts = daycuts / 365)
fit.ph1 <- tpchreg(oe(event, exposure) ~ 1, data = bb, time = age)
plot(fit.ph1, fn = "haz", axes = FALSE, xlab = "Days", main = "", lwd = 1.5, col = "blue")
dax <- c(0, 28, 91, 183, 365)
axis(1, at = dax / 365, labels = dax)
axis(2, las = 1)
box()
abline(v = 28 / 365, lty = 3)
abline( h = 0, v = 0)
```

A calmer development for the postneonatal mortality, see Figure \@ref(fig:phneo).

```{r phneo, fig.cap = "Crude postneonatal mortality."}
b2 <- age.window(b1, c(28 / 365, 1))
pcuts <- c(28, seq(49, 245, by = 28), 285, 324, 365)
tp <- toTpch(Surv(enter, exit, event) ~ 1, data = b2, cuts = pcuts / 365)

fit.ph <- tpchreg(oe(event, exposure) ~ 1, data = tp, time = age)
plot(fit.ph, fn = "haz", axes = FALSE, xlab = "Days", main = "", lwd = 1.5, col = "blue")
axis(1, at = pcuts / 365, labels = pcuts)
axis(2, las = 1)
box()
##abline(v = 28 / 365, lty = 3)
abline( h = 0, v = 0)
```

Figure \@ref(fig:phneo) suggests that some smooth parametric model could fit well,
for instance the *left truncated Weibull*  distribution. Let's see (Figure \@ref(fig:weib)).

```{r weib, fig.cap = "A perfect Weibull hazard fit."}
##births <- readRDS("data/births4.rds")
postbirths <- age.window(births, c(28, 365))
postbirths$exit <- postbirths$exit / 365
postbirths$enter <- postbirths$enter / 365
fit <- phreg(Surv(enter, exit, event) ~ 1, data = postbirths)
plot(fit, fn = "haz", lwd = 1.5, col = "blue", main = "", xlab = "Days", axes = FALSE)
axis(1, at = dax / 365, labels = dax)
axis(2, las = 1)
abline(h = 0, v = 28 / 365)
box()
hh <- hazards(fit.ph, cum = FALSE)
lines(hh$x, hh$y, col = "red")
legend("topright", legend = c("Truncated Weibull", "Piecewise constant"), col = c("blue", "red"), lwd = c(1.5,1))
```

A perfect fit!

In the following we split the analyses into the neonatal and the postneonatal periods, with
a focus on the latter. But we start with a short comment on the link to *biometric* analysis
à la @bp51a.

## The biometric finding

What happens if we extend the truncated Weibull back into neonatal territory? The answer
is given in Figure \@ref(fig:neowei).

```{r neowei, fig.cap = "Weibull hazard functions, total and exogenous neonatal mortality.", fig.height = 4}
b3 <- age.window(b1, c(0, 28 / 365))
fit.w0 <- phreg(Surv(enter, exit, event) ~ 1, data = b3)
plot(fit.w0, fn = "haz", axes = FALSE, xlab = "Days", main = "", lwd = 1.5, col = "red", ylim = c(0, 3))
#fit.ph <- tpchreg(oe(event, exposure) ~ 1, data = inftab, time = age)
##plot(fit.ph1, fn = "haz", axes = FALSE, xlab = "Days", main = "", lwd = 1.5, col = "red",
  ##   xlim = c(0, 28 / 365))
dor <- seq(0, 28, by = 2)
axis(1, at = dor / 365, labels = dor)
axis(2, las = 1)
box()
##abline(v = 28 / 365, lty = 3)
abline( h = 0, v = 0)
pars <- exp(fit$coefficients)
x <- seq(0, 28/365, length = 1000)
y <- hweibull(x, shape = pars[2], scale = pars[1])
lines(x, y, col = "blue", lwd = 1.5)
legend("topright", legend = c("Weibull postneonatal (exogenous), extrapolated", "Weibull neonatal (total)"), col = c("blue", "red"), lwd = c(1.5, 1))
```

Not very good fit in this territory. The difference here is that to the exogenous mortality, endogenous mortality is added.
The difference between the piecewise constant and the Weibull in the figure would be 
called the *endogeneous infant mortality* [@bp51a; @bp51b].

@@

The neonatal mortality in our geographical area and time period is thoroughly analysed
in two papers by @karlsson and @junkka and we only present a brief confirmatory analysis.

No strong signs of interactions with marginTemp and expTemp data, so these factors strike
almost evenly over different subpopulations. Moreover, the effects are stronger during winter,
with a decrease in mortality for each degree temperature increase, almost the same amount 
for both, around 1.2 per cent.

Separate analyses for winter and summer, see Tables \@ref(tab:neowinter) and 
\@ref(tab:neosummer) follow.

@@

```{r neowinter, results = "asis"}
##neo$marginTemp <- neo$extemp.1
##neo$expTemp <- neo$emeantemp.1
fit.neow <- coxreg(Surv(enter, exit, event) ~ expTemp + marginTemp + socst + 
                      sex + urban + parity + I(year - 1920),
                  data = neo[!neo$summer, ])
drw <- drop1(fit.neow, test = "Chisq")
ltx(fit.neow, dr = drw, caption = "Neonatal mortality, winter.", label = "tab:neowinter")
```

The result of a Cox regression of neonatal mortality during winter is shown in Table \@ref(tab:neowinter).
There is a (statistically) strong effect of *expTemp*, an increase of the expected temperature
lowers neonatal mortality by around 1.3 per cent per degree. Almost the same effect is shown for *marginTemp*.

The interaction between social status and expTemp is shown in Figure \@ref(fig:plotwcsn). Nothing interesting is
going on, the effect of expTemp on neonatal mortality is the same in all social classes ("no interaction" between
expTemp and socst).

```{r plotwcsn, fig.cap = "Socioeconomic status and expTemp, winter. Neonatal mortality.", fig.height = 3}
source("R/socfig.R")
winter <- neo[!neo$summer, ]
fitw <- coxreg(Surv(enter, exit, event) ~ socst:expTemp + socst, data = winter)
fitw0 <- coxreg(Surv(enter, exit, event) ~ socst + expTemp, data = winter)
p <- anova(fitw, fitw0)[[4]][2]
rang <- range(winter$expTemp)
socfig(coef(fitw), main = "", pval = p, ran = rang, ymax = 4)
```

The interaction between social status and marginTemp is shown in Figure \@ref(fig:plotwwsn). A little
bit more action here, but the conclusion stays the same: No interaction between marginTemp and social class,
bad marginTemp hits all by the same strength.

```{r plotwwsn, fig.cap = "Socioeconomic status and marginTemp, winter. Neonatal mortality.", fig.height = 3}
source("R/socfig.R")
fitww <- coxreg(Surv(enter, exit, event) ~ socst:marginTemp + socst, data = winter)
fitww0 <- coxreg(Surv(enter, exit, event) ~ socst + marginTemp, data = winter)
p <- anova(fitww, fitww0)[[4]][2]
rang <- range(winter$marginTemp)
socfig(coef(fitww), main = "", pval = p, ran = rang, ymax = 4)
```

@@

The summer result of a Cox regression is shown in Table \@ref(tab:neosummer).
Compared to winter, the *expTemp* effect is in the same direction but weaker, while 
the effect of *marginTemp* is zero.

```{r neosummer, results = "asis"}
##neo$marginTemp <- neo$extemp.1
##neo$expTemp <- neo$emeantemp.1
fit.neos <- coxreg(Surv(enter, exit, event) ~ expTemp + marginTemp + socst + 
                      sex + urban + parity + I(year - 1920),
                  data = neo[neo$summer, ])
##data = neo[neo$week %in% 20:32, ])
drs <- drop1(fit.neos, test = "Chisq")
ltx(fit.neos, dr = drs, caption = "Neonatal mortality, summer.", label = "tab:neosummer")
```

During summer, expTemp and marginTemp effects are clearly smaller, and regarding marginTemp, the opposite 
direction: Excess heat increases the death risk, however not very much and not statistically significant.
